### 2014-02-20 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd
h

This document has three main parts
- Reproduce some of the results to show that we are working on the same data and analysis workflow.
- Remove the use of ComBat and perform the same analysis with alternative established tools and estimate the error introduced by ComBat.
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is an error introduced by ComBat.

@@

orgdatamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
datamatrix=orgdatamatrix[1:3000,]# for dev.
#datamatrix=orgdatamatrix# for dev.
sampleannotation = read.table("data/sampleInfoExample2.txt", 
                              sep="\t", header=TRUE, stringsAsFactors=FALSE)
rownames(sampleannotation)=sampleannotation$ArrayName
sampleannotation$Batch=factor(as.character(sampleannotation$Batch)) 
sampleannotation$Cell=factor(as.character(sampleannotation$Cell)) 
# must be discrete for the pheatmap
#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]]) # ordercheck
datamatrix=datamatrix[,sampleannotation$Type!="WT"]
sampleannotation=sampleannotation[sampleannotation$Type!="WT",]
sampleannotation$Type=factor(sampleannotation$Type)
print(dim(datamatrix))
#dev/debug
useparprior=TRUE
```
The covariate/batch balance;
```{r}
print(table(sampleannotation$Type, sampleannotation$Batch))
```

@@

Result when including batch in the statistical test
--------------

An alternative (and better?) way of handling batch effect is to include it in the model for the statistical test. This is possible in several tools and we choose to use the popular limma package (Smyth et al). Since this is not the tool used above we will first create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
```{r}
Type = as.factor(sampleannotation$Type)
design = model.matrix(~0 + Type)
fit <- lmFit(EB3, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
EB3_limma_pvalues = fit2$p.value[,1]
print(table(qvalue(EB3_limma_pvalues)$qvalue<0.05))
```
The number of genes below the significance threshold is fairly similar to the t-test.

Next the same test is applied to the data whithout ComBat adjustment, but with the batch included as a blocking factor.
```{r}
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design <- model.matrix(~0+Type+Block)
fit <- lmFit(datamatrix, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
datamatrix_limma_pvalues = fit2$p.value[,1]
print(table(qvalue(datamatrix_limma_pvalues)$qvalue<0.05))
```
Which gives a somwhat lower number.

Now we can compare the P-values for the two alternative procedures.
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EB3_limma_pvalues, border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
hist(datamatrix_limma_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted"), text.col=c("blue", "red"))
```
This plot shows that the p-values calculated from the ComBat adjusted data is skewed towards the low end compared to when batch is considered inside the statistical test in limma. However, the difference is not large and from this plot alone it could be argued that ComBat is just better.

Lets assume the aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05. Then handling the batch effect with ComBat adjustment will yield a list of `r sum(qvalue(EB3_limma_pvalues)$qvalue<0.05)` wheras a list of `r sum(qvalue(datamatrix_limma_pvalues)$qvalue<0.05)` probes would have been closer to the truth.

table( qvalue(datamatrix_limma_pvalues)$qvalue<0.05 & qvalue(EB3_limma_pvalues)$qvalue<0.05)

Additional sanity checks
---------

```{r, results='hide', tidy=FALSE}
set.seed(100)
randommatrix = datamatrix
randommatrix[,] =rnorm(length(datamatrix), mean=0, sd=1)
EBrand = as.matrix( sva::ComBat(
            dat=randommatrix, 
            batch=sampleannotation[colnames(randommatrix),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(randommatrix),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
```
```{r, tidy=FALSE}

@@

```

```{r, dev='svg', fig.width=8, fig.height=8}
hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are enriched slighlty for low p-values. Indicating that the same enrichment seen for the real data is also false.

>>>>

This document has four main parts
- Reproduce some of the results to show that we are working on the same data and analysis workflow.
- Remove the use of ComBat and perform the same analysis with alternative established tools
- Estimate the error introduced by ComBat and its consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is an error introduced by ComBat.

@@

orgdatamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
datamatrix=orgdatamatrix[1:3000,]# for dev.
#datamatrix=orgdatamatrix# for dev.
sampleannotation = read.table("data/sampleInfoExample2.txt", 
                              sep="\t", header=TRUE, stringsAsFactors=FALSE)
rownames(sampleannotation)=sampleannotation$ArrayName
sampleannotation$Batch=factor(as.character(sampleannotation$Batch)) 
sampleannotation$Cell=factor(as.character(sampleannotation$Cell)) 
# must be discrete for the pheatmap
#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]]) # ordercheck
datamatrix=datamatrix[,sampleannotation$Type!="WT"]
sampleannotation=sampleannotation[sampleannotation$Type!="WT",]
sampleannotation$Type=factor(sampleannotation$Type)
print(dim(datamatrix))
#dev/debug
useparprior=TRUE
```
The covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```

@@

Analysis without ComBat
--------------

An alternative (and better?) way of handling batch effect is to include it in the model for the statistical test. This is possible in several tools and we choose to use the popular limma package (Smyth et al).

We start with the data before ComBat adjustment. Limma works best with log transformed and between array normalized values, so first we have to do something with the negative signals.  Then the test is run with the batch included as a blocking factor.
```{r}
thismat = datamatrix
thismat[thismat<1]=1
thismat = log2(thismat)
thismat=normalizeBetweenArrays(thismat)
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design <- model.matrix(~0+Type+Block)
fit <- lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
datamatrix_limma_res <- eBayes(fit2)
print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
```

Consequences of ComBat use for the end result
--------------

Since limma was not the tool used in Johnson et al.s workflow, we will first create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
```{r}
thismat = EB3
thismat[thismat<1]=1
thismat = log2(thismat)
thismat=normalizeBetweenArrays(thismat)
Type = as.factor(sampleannotation$Type)
design = model.matrix(~0 + Type)
fit <- lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
EB3_limma_res <- eBayes(fit2)
print(table(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05))
```
The number of genes below the significance threshold is fairly similar to the t-test.

Now we can compare the P-values for the two alternative procedures.
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EB3_limma_res$p.value[,1], border="blue", main="P-values, ComBat vs Limma", breaks=100, xlab="p-value")
hist(datamatrix_limma_res$p.value[,1], border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted"), text.col=c("blue", "red"))
```
This plot shows that the p-values calculated from the ComBat adjusted data is skewed towards the low end compared to when batch is considered inside the statistical test in limma. However, the difference is not large and from this plot alone it could be argued that ComBat is just better.

Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05. Then handling the batch effect with ComBat adjustment will yield a list of `r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)` wheras a list of `r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)` probes would have been closer to the truth. Also it is of interest to note that it is mostly the same probes that are found.
```{r, tidy=FALSE}
table( qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05 ,
         qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05,
       dnn=c("limma", "ComBat"))
```
And if they were to use the top 1000 probes in a gene set test they found have used many of the same genes regardless of the use of ComBat.
```{r, tidy=FALSE}
table( rank(datamatrix_limma_res$p.value[,1])  <=1000 & 
         rank(EB3_limma_res$p.value[,1])<=1000   )
```
Our conclusion is that for this study the error introduced by the use of ComBat would only have a modest effect on the final result.

Additional sanity checks
---------
For this data set ComBat's handling of batch-effects yields only modestly different results compared to the alternative. Explanation for the difference could be that ComBat is better. To substantiate that the result from the use of ComBat is less trustworthy we provide a few additional sanity checks.

First we use random numbers drawn from the same distribution regardless of batch or covariate but retain the batch/covariate design.
```{r, results='hide', tidy=FALSE}
set.seed(100)
randommatrix = datamatrix
randommatrix[,] =rnorm(length(datamatrix), mean=0, sd=1)
EBrand = as.matrix( sva::ComBat(
            dat=randommatrix, 
            batch=sampleannotation[colnames(randommatrix),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(randommatrix),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
```
Limma is then used in 3 ways
- On the ComBat adjusted random numbers
- On the random numbers including batch as a blocking factor
- On the random numbers ignorant to batch information
```{r, tidy=FALSE}

@@

```
Plotting the p-values
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are enriched slighlty for low p-values. Indicating that the same enrichment seen for the real data is also false.
